---
title: "P8105 HW3"
author: "Elaine Yanxi Chen"
date: "`r Sys.Date()`"
output: github_document
---

## Packages and settings

First we load the packages necessary to knit this document.

```{r packages and settings, message = FALSE}
library(tidyverse)
library(ggridges)
library(patchwork)

library(p8105.datasets)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


# Problem 1: Instacart data

## Import data

```{r load instacart}
library(p8105.datasets)
data("instacart")
```

## Describe the dataset

* This dataset contains `r nrow(instacart)` rows and `r ncol(instacart)` columns, with each row representing a single product from an instacart order.

* Variables include identifiers for user, order, and product; the order in which each product was added to the cart. There are several order-level variables, describing the day and time of the order, and number of days since prior order. Then there are several item-specific variables, describing the product name (e.g. Yogurt, Avocado), department (e.g. dairy and eggs, produce), and aisle (e.g. yogurt, fresh fruits), and whether the item has been ordered by this user in the past. 

* In total, there are `r instacart %>% select(product_id) %>% distinct %>% count` products found in `r instacart %>% select(user_id, order_id) %>% distinct %>% count` orders from `r instacart %>% select(user_id) %>% distinct %>% count` distinct users.

## Answer questions

* How many aisles are there, and which aisles are the most items ordered from?

```{r}
instacart %>% 
  count(aisle) %>% 
  arrange(desc(n))
```

There are `r instacart %>% select(aisle_id) %>% distinct %>% count` aisles, and the most items are ordered from the fresh vegetables and fresh fruits aisles. 